Analysis of Human Activity Pattern
========================================================
#### Andrey Alferov
#### Sunday, February 15, 2015

### Load required libraries
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
```

### Download dataset file, unzip it and load in dataframe
```{r, warning=FALSE, message=FALSE}
if(!file.exists("./data"))
{dir.create("./data")}

url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
fileName <- "repdata_Fdata_Factivity.zip"

download.file(url, destfile = fileName)

unzip(fileName, exdir="data")

df <- read.csv(file="./data/activity.csv",head=TRUE,sep=",")
df$date <- as.Date(df$date)

summary(df)
```


## Analize what is mean total number of steps taken per day
### Calculate the total number of steps taken per day
```{r}
daily <- group_by(na.omit(df), date)
per_day <- summarise(daily, steps = sum(steps))
summary(per_day)
```
### Make a histogram of the total number of steps taken each day
```{r, fig.width=10}
plot <- qplot(as.POSIXct(date), steps, data=per_day, geom="histogram", stat="identity")
plot + xlab("Date") + ylab("Number of steps per day") + scale_x_datetime(breaks = date_breaks("1 week"))
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r}
summarise(per_day, mean = mean(steps), median = median(steps))
```

## Analize what is the average daily activity pattern
### Calculate the average number of steps taken per 5-minute interval across all days
```{r}
by_interval <- group_by(na.omit(df), interval)
per_interval <- summarise(by_interval, steps = mean(steps))
summary(per_interval)
```

### Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
ggplot(per_interval, aes(interval, steps)) + geom_line() + xlab("Time interval") + ylab("Mean number of steps") +
       scale_x_discrete(breaks = c(0,600,1200,1800,2400), labels = c("0:00","6:00","12:00","18:00","24:00"))
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_steps <- summarise(per_interval, max_steps = max(steps))
filter(per_interval, steps == as.numeric(max_steps))
```
The maximum number of steps occurs in the 5-minutes interval of the day starting at 8:35.

## Imputing missing values
### Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(df$steps)) 
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
df_new <- df

for (i in 1:nrow(df_new)){
  if (is.na(df_new$steps[i])){
    df_new$steps[i] <- per_interval$steps[which(df_new$interval[i] == per_interval$interval)]}
}

df_new <- arrange(df_new, interval)
summary(df_new)
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
```{r, fig.width=10}
daily_new <- group_by(df_new, date)
per_day_new <- summarise(daily_new, steps = sum(steps))
plot <- qplot(as.POSIXct(date), steps, data=per_day_new, geom="histogram", stat="identity")
plot + xlab("Date") + ylab("Number of steps per day") + scale_x_datetime(breaks = date_breaks("1 week"))
```
### Calculate and report the mean and median of the total number of steps taken per day
```{r}
summarise(per_day_new, mean = mean(steps), median = median(steps))
```
The median value of the total number of steps taken per day have slightely increased compared to the original data set and became the same as the mean value.

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
difference <- sum(df_new$steps) - sum(na.omit(df)$steps)
round(difference)
```
Due to imputation the total sum of steps increased by 86130 steps.

## Analize if there are differences in activity patterns between weekdays and weekends
```{r}
Sys.setlocale("LC_TIME", "English")
```

### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
```{r}
df_new$weekday <- factor(weekdays(df_new$date) %in% c("Saturday","Sunday"), 
                                  labels=c("weekday","weekend"), ordered=FALSE)
```

### Calculate the average number of steps taken per 5-minute interval across all days
```{r}
by_interval <- group_by(df_new, interval, weekday)
per_interval <- summarise(by_interval, steps = mean(steps))
summary(per_interval)
```

### Make a time series plot
```{r}
ggplot(per_interval, aes(interval, steps)) + geom_line() + facet_grid(weekday ~ .) + 
  xlab("Time interval") + ylab("Mean number of steps") +
  scale_x_discrete(breaks = c(0,600,1200,1800,2400), labels = c("0:00","6:00","12:00","18:00","24:00"))
```



